A Dataset of Scratch Programs:\\Scraped, Shaped and Scored

### 2017-03-16 MSR-Scratch.tex
Minor changes. For me, it is ready to be submitted.

\author{
	\IEEEauthorblockN{Efthimia Aivaloglou\IEEEauthorrefmark{1}, Felienne Hermans\IEEEauthorrefmark{1}, Jes{\'u}s Moreno-Le{\'o}n\IEEEauthorrefmark{2}, and Gregorio Robles\IEEEauthorrefmark{3}}
	\IEEEauthorblockA{\IEEEauthorrefmark{1}Delft University of Technology, The Netherlands\\
			Email: \{e.aivaloglou, f.f.j.hermans\}@tudelft.nl}
	\IEEEauthorblockA{\IEEEauthorrefmark{2}Programamos.es \& Universidad Rey Juan Carlos, Spain\\
		Email: jesus.moreno@programamos.es}
	\IEEEauthorblockA{\IEEEauthorrefmark{3}Universidad Rey Juan Carlos, Fuenlabrada (Madrid), Spain \\
		Email: grex@gsyc.urjc.es}
}

@@

A number of works in the computing education research field attempt to assess the programming skills that novice programmers develop in the Scratch environment.
Some utilize program data collected during specific programming courses (e.g.,~\cite{meerbaum-salant_learning_2010, wilson_evaluation_2012, Maloney_2008}), while others utilize the dataset made available by the Lifelong Kindergarten Group at the MIT Media Lab~\cite{2017_scratch_dataset}, which contains data for Scratch projects created until 2012 (e.g.,~\cite{fields_2014, yang_2015, Dasgupta_2016}).
In addition to identifying indications of learning of programming concepts, static analysis of Scratch programs has also been performed for identifying code smells and bad programming practices ~\cite{Meerbaum_habits_2011, Aivaloglou_2016}, and automated quality assessment tools have been proposed (e.g., Hairball~\cite{boe_hairball:_2013} and Dr. Scratch~\cite{moreno_automatic_2014}).

While Scratch is receiving increasing interest as an introductory programming language, there is no recent dataset of Scratch programs available to the research community.
The one made available from the MIT Media Lab concerns projects created using the previous, initial version of the Scratch application, before the introduction of the current Scratch version (Scratch 2) and of the web programming interface in 2013.
It is since then that the popularity of Scratch started to increase.\footnote{Monthly activity trends can be found at \url{https://scratch.mit.edu/statistics/}}
Scratch 2 introduced several features relating to important programming concepts,\footnote{The complete list of features introduced in Scratch 2 can be found at \url{https://wiki.scratch.mit.edu/wiki/Scratch_2.0}} like custom blocks (the equivalent to procedure definitions), and therefore that dataset does not include projects utilizing those features.

@@

We ran the scraper on March 2nd 2016 for 24 hours and, during that time, it obtained the JSON files for \nPrograms~projects. Out of those, we failed to parse and further analyze 2,367 projects due to technical difficulties with the JSON files.
The scraper and all project files are available\footnotemark[\ref{dataseturl}].

@@

\begin{table}[]
	\centering
	\begin{tabular}{llp{5.4cm}}
		\textbf{Table}& \textbf{Key} & \textbf{Attribute(Description)}\\
		\hline
		\textbf{\texttt{projects}} & PK & \texttt{p\_id} (Scratch project ID)\\
		&  & 							\texttt{project-name} (name given to project)\\
		&  & 							\texttt{username} (author's Scratch 
		username)\\
		&  & 							\texttt{total-views} (project views number)\\
		&  & 							\texttt{total-remixes} (project remixes number)\\
		&  & 							\texttt{total-favorites} (total users favoriting)\\
		&  & 							\texttt{total-loves} (users `loving' the project)\\
		& & 							\texttt{is-remix} (calculated column, true if project is a remix of another one)\\
		\hline
		\textbf{\texttt{scripts}} & PK & \texttt{script\_id} (auto increment)\\
		& FK & 							\texttt{project\_id} (\texttt{\scriptsize{*\ldots1 projects:p\_id}})\\
		& & 							\texttt{sprite-type} ([$sprite \vert stage \vert procDef$])\\
		& & 							\texttt{sprite-name} (name of sprite the script is in)\\
		& & 							\texttt{script-rank} (project-level ranking of script)\\
		& & 							\texttt{coordinates} (x-y location of script in Scratch editor)\\
		& & 							\texttt{total-blocks} (calculated column, number of blocks comprising the script)\\
		\hline
		\textbf{\texttt{procedures}} & PK,FK & \texttt{script\_id} (\texttt{\scriptsize{1\ldots1 scripts:script\_id}})\\
		& & 							\texttt{proc-name} (name given to custom block)\\
		& & 							\texttt{total-args} (number of input arguments)\\
		\hline
		\textbf{\texttt{blocks}} & PK,FK  & \texttt{script\_id} (\texttt{\scriptsize{*\ldots1 scripts:script\_id}})\\
		& PK & \texttt{block-rank} (script-level ranking of block)\\
		&  & \texttt{block-type} (\texttt{\scriptsize{*\ldots1 blockTypes:b-type}})\\
		&  & \texttt{parameter-1} (value of 1st input parameter)\\
		&  & ... \\
		&  & \texttt{parameter-24} (value 24th input parameter)\\
		\hline
		\textbf{\texttt{blockTypes}} & PK & \texttt{b-type} (Scratch name of predefined block)\\
		&  & \texttt{category} (Scratch block category)\\
		&  & \texttt{shape} (One of 8 block shapes in Scratch)\\
		&  & \texttt{is-input} (true if block receives user input)\\
		\hline
		\textbf{\texttt{remixes}} & PK  & \texttt{remix\_id} (\texttt{\scriptsize{project\_id}} of created project)\\
		&  & \texttt{from-project\_id} (\texttt{\scriptsize{project\_id}} of original project)\\
		\hline
		\textbf{\texttt{grades}} & PK,FK & \texttt{project\_id} (\texttt{\scriptsize{1\ldots1 projects:p\_id}})\\
% Original
%		& & \texttt{Abstraction} (Score for abstraction and problem decomposition)\\
%		& & \texttt{Parallelism} (Score for parallelism)\\
%		& & \texttt{Logic} (Score for logical thinking)\\
%		& & \texttt{Synchronization} (Score for synchronization)\\
%		& & \texttt{FlowControl} (Score for algorithmic notions of flow control)\\
%		& & \texttt{UserInteractivity} (Score for user interactivity)\\
%		& & \texttt{DataRepresentation} (Score for data representation)\\

@@

The schema of the database, outlined in Table~\ref{tbl-dbschema}, reflects the structure of the Scratch programs.
Projects contain sprites, which are entities with their own associated code.
The code is organized into \textbf{\texttt{scripts}}, i.e., groups of Scratch code blocks, each script belonging to a sprite named \texttt{sprite-name}.
In the example in Figure~\ref{fig-scratchExample} there are 3 scripts, one initiated when the green flag is clicked, one then the sprite is clicked, and the `backflip' custom block definition.
Those custom blocks are the equivalent of procedure definitions, with their names and arguments stored in the \textbf{\texttt{procedures}} table.

@@

Table~\ref{tbl-stats} also reveals the median project in the dataset: it contains 2 sprites, 4 scripts, 26 blocks, no custom block definitions, and receives a mastery score of 8 our of 21, which is considered a medium level of computational thinking skills development. The table highlights the existence of surprisingly complex projects, with 525 sprites and over 34,000 blocks.

@@

Apart from quantitative studies, the dataset can support the design of experiments and field studies, whose material often includes Scratch programs with specific characteristics.
For example, in~\cite{hermans_2016} examples of Scratch programs exhibiting specific smells were used in a controlled experiment to determine how the smells affect the understanding and the modification of the programs.
This queryable dataset can support finding example programs with characteristics according to the experiment design and the field of the study.

\section{Limitations}
The only information about the authors of the Scratch projects contained in the dataset is their username.
This suffices for creating project portfolios of the authors and for facilitating research in the directions described in the previous section.
However, other directions in the computing education research field require richer author data, and especially data on their gender and age.
There are, for example, indications that specific programming concepts are better understood after certain ages ---Seiter and Foreman~\cite{Seiter_2013} analyzed 150 Scratch projects from primary school students and found that design patterns requiring the understanding of parallelism, conditionals and, especially, variables were under-represented by all grades apart from 5 an 6.
The effect of gender and account age were also examined in~\cite{fields_2014} in relation to the use of programming concepts.
%Studies on the effect of parameters like the age are not only interesting but also of direct value to educators, since they give insights on when it is better to introduce computational thinking concepts in the curriculum of schools.
However, this dataset does not include information on the gender and the age of authors and cannot be extended to do so using the current Scratch web interface because this information is not available in the user profile page\footnotemark[\ref{fn-authorpage}], even though it is provided by users upon registration.

>>>>

\author{
	\IEEEauthorblockN{Efthimia Aivaloglou\IEEEauthorrefmark{1}, Felienne Hermans\IEEEauthorrefmark{1}, Jes{\'u}s Moreno-Le{\'o}n\IEEEauthorrefmark{2}, and Gregorio Robles\IEEEauthorrefmark{3}}
	\IEEEauthorblockA{\IEEEauthorrefmark{1}Delft University of Technology, The Netherlands\\
			\{e.aivaloglou, f.f.j.hermans\}@tudelft.nl}
	\IEEEauthorblockA{\IEEEauthorrefmark{2}Programamos.es \& Universidad Rey Juan Carlos, Spain\\
		jesus.moreno@programamos.es}
	\IEEEauthorblockA{\IEEEauthorrefmark{3}Universidad Rey Juan Carlos, Fuenlabrada (Madrid), Spain \\
		grex@gsyc.urjc.es}
}

@@

A number of works in the computing education research field attempt to assess the programming skills that novice programmers develop in the Scratch environment.
Some utilize program data collected during specific programming courses (e.g.,~\cite{meerbaum-salant_learning_2010, wilson_evaluation_2012, Maloney_2008}), while others utilize the dataset made available by the Lifelong Kindergarten Group at the MIT Media Lab~\cite{2017_scratch_dataset}, which contains data for Scratch projects created until 2012 (e.g.,~\cite{fields_2014, yang_2015, Dasgupta_2016}).
In addition to identifying indications of learning of programming concepts, static analysis of Scratch programs has also been performed for identifying code smells and bad programming practices (e.g.,~\cite{Meerbaum_habits_2011, Aivaloglou_2016}), and automated quality assessment tools have been proposed (e.g., Hairball~\cite{boe_hairball:_2013} and Dr. Scratch~\cite{moreno_automatic_2014}).

While Scratch is receiving increasing interest as an introductory programming language, there is no recent dataset of Scratch programs available to the research community.
The one made available from the MIT Media Lab concerns projects created using the previous, initial version of the Scratch application, before the introduction of the current Scratch version (Scratch 2) and of the web programming interface in 2013.
It is since then that the popularity of Scratch started to increase\footnote{Monthly activity trends can be found at \url{https://scratch.mit.edu/statistics/}}.
Scratch 2 introduced several features relating to important programming concepts\footnote{The complete list of features introduced in Scratch 2 can be found at \url{https://wiki.scratch.mit.edu/wiki/Scratch_2.0}}, like custom blocks (the equivalent to procedure definitions), and therefore that dataset does not include projects utilizing those features.

@@

We ran the scraper on March 2\textsuperscript{nd} 2016 for 24 hours and, during that time, it obtained the JSON files for \nPrograms~projects. Out of those, we failed to parse and further analyze 2,367 projects due to technical difficulties with the JSON files.
The scraper and all project files are available\footnotemark[\ref{dataseturl}].

@@

\begin{table}[]
	\centering
	\begin{tabular}{llp{5.4cm}}
		\textbf{Table}& \textbf{Key} & \textbf{Attribute(Description)}\\
		\hline
		\textbf{\texttt{projects}} & PK & \texttt{p\_id} (Scratch project ID)\\
		&  & 							\texttt{project-name} (name given to project)\\
		&  & 							\texttt{username} (author's Scratch 
		username)\\
		&  & 							\texttt{total-views} (project views number)\\
		&  & 							\texttt{total-remixes} (project remixes number)\\
		&  & 							\texttt{total-favorites} (total users favoriting)\\
		&  & 							\texttt{total-loves} (users `loving' the project)\\
		& & 							\texttt{is-remix} (calculated column, true if project is a remix of another one)\\
		\hline
		\textbf{\texttt{scripts}} & PK & \texttt{script\_id} (auto increment)\\
		& FK & 							\texttt{project\_id} (\texttt{\scriptsize{*\ldots1 projects:p\_id}})\\
		& & 							\texttt{sprite-type} ([$sprite \vert stage \vert procDef$])\\
		& & 							\texttt{sprite-name} (name of sprite the script is in)\\
		& & 							\texttt{script-rank} (project-level ranking of script)\\
		& & 							\texttt{coordinates} (x-y location of script in Scratch editor)\\
		& & 							\texttt{total-blocks} (calculated column, number of blocks comprising the script)\\
		\hline
		\textbf{\texttt{procedures}} & PK,FK & \texttt{script\_id} (\texttt{\scriptsize{1\ldots1 scripts:script\_id}})\\
		& & 							\texttt{proc-name} (name given to custom block)\\
		& & 							\texttt{total-args} (number of input arguments)\\
		\hline
		\textbf{\texttt{blocks}} & PK,FK  & \texttt{script\_id} (\texttt{\scriptsize{*\ldots1 scripts:script\_id}})\\
		& PK & \texttt{block-rank} (script-level ranking of block)\\
		&  & \texttt{block-type} (\texttt{\scriptsize{*\ldots1 blockTypes:b-type}})\\
		&  & \texttt{parameter-1} (value of 1\textsuperscript{st} input parameter)\\
		&  & ... \\
		&  & \texttt{parameter-24} (value 24\textsuperscript{th} input parameter)\\
		\hline
		\textbf{\texttt{blockTypes}} & PK & \texttt{b-type} (Scratch name of predefined block)\\
		&  & \texttt{category} (Scratch block category)\\
		&  & \texttt{shape} (One of 8 block shapes in Scratch)\\
		&  & \texttt{is-input} (true if block receives user input)\\
		\hline
		\textbf{\texttt{remixes}} & PK  & \texttt{remix\_id} (\texttt{\scriptsize{project\_id}} of created project)\\
		&  & \texttt{from-project\_id} (\texttt{\scriptsize{project\_id}} of original project)\\
		\hline
		\textbf{\texttt{grades}} & PK,FK & \texttt{project\_id} (\texttt{\scriptsize{1\ldots1 projects:p\_id}})\\
% Original
%		& & \texttt{Abstraction} (Score for abstraction and problem decomposition)\\
%		& & \texttt{Parallelism} (Score for parallelism)\\
%		& & \texttt{Logic} (Score for logical thinking)\\
%		& & \texttt{Synchronization} (Score for synchronization)\\
%		& & \texttt{FlowControl} (Score for algorithmic notions of flow control)\\
%		& & \texttt{UserInteractivity} (Score for user interactivity)\\
%		& & \texttt{DataRepresentation} (Score for data representation)\\

@@

The schema of the database, outlined in Table~\ref{tbl-dbschema}, reflects the structure of the Scratch programs.
Projects contain sprites, which are entities with their own associated code.
The code is organized into \textbf{\texttt{scripts}}, i.e., groups of Scratch code blocks, each script belonging to a sprite named \texttt{sprite-name}.
In the example in Figure~\ref{fig-scratchExample} there are three scripts, one initiated when the green flag is clicked, one then the sprite is clicked, and the `backflip' custom block definition.
Those custom blocks are the equivalent of procedure definitions, with their names and arguments stored in the \textbf{\texttt{procedures}} table.

@@

Table~\ref{tbl-stats} also reveals the median project in the dataset: it contains 2 sprites, 4 scripts, 26 blocks, no custom block definitions, and receives a mastery score of 8 out of 21, which is considered a medium level of computational thinking skills development. The table highlights the existence of surprisingly complex projects, with 525 sprites and over 34,000 blocks.

@@

Apart from quantitative studies, the dataset can support the design of experiments and field studies, whose material often includes Scratch programs with specific characteristics.
For example, examples of Scratch programs exhibiting specific smells have been used in a controlled experiment to determine how the smells affect the understanding and the modification of the programs~\cite{hermans_2016}.
This queryable dataset can support finding example programs with characteristics according to the experiment design and the field of the study.

\section{Limitations}
The only information about the authors of the Scratch projects contained in the dataset is their username.
This suffices for creating project portfolios of the authors and for facilitating research in the directions described in the previous section.
However, other directions in the computing education research field require richer author data, and especially data on their gender and age.
There are, for example, indications that specific programming concepts are better understood after certain ages ---Seiter and Foreman~\cite{Seiter_2013} analyzed 150 Scratch projects from primary school students and found that design patterns requiring the understanding of parallelism, conditionals and, especially, variables were under-represented by all grades apart from 5\textsuperscript{th} and 6\textsuperscript{th}.
The effect of gender and account age were also examined in~\cite{fields_2014} in relation to the use of programming concepts.
%Studies on the effect of parameters like the age are not only interesting but also of direct value to educators, since they give insights on when it is better to introduce computational thinking concepts in the curriculum of schools.
However, this dataset does not include information on the gender and the age of authors, and cannot be extended to do so using the current Scratch web interface because this information is not available in the user profile page\footnotemark[\ref{fn-authorpage}], even though it is provided by users upon registration.
